---
title: "Simple"
author: "DUO Hongrui"
output:
    BiocStyle::html_document:
        toc: true
        toc_float: true
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Simulation method: Simple}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---
<!-- github markdown built using 
rmarkdown::render("vignettes/Simple.Rmd", output_format = rmarkdown::github_document())
-->

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Here Simple method will be demonstrated clearly and hope that this document can help you.

# Estimating parameters from a real dataset

Before simulating datasets, it is important to estimate some essential parameters from a real dataset in order to make the simulated data more real. If you do not have a single-cell transcriptomics count matrix now, you can use the data collected in simmethods package by `simmethods:data` command.

```{r load_data, message=FALSE, warning=FALSE, error=FALSE}
library(simmethods)
library(SingleCellExperiment)
# Load data
ref_data <- simmethods::data
dim(ref_data)
```

Using `simmethods::Simple_estimation` command to execute the estimation step.

```{r estimate}
estimate_result <- simmethods::Simple_estimation(ref_data = ref_data,
                                                 verbose = TRUE,
                                                 seed = 111)
```

# Simulating datasets using Simple

After estimating parameter from a real dataset, we will simulate a dataset based on the learned parameters with different scenarios.

1. Datasets with default parameters
2. Determin the number of cells and genes

## Datasets with default parameters

The reference data contains 160 cells and 4000 genes, if we simulate datasets with default parameters and then we will obtain a new data which has the same size as the reference data.

```{r default_params}
simulate_result <- simmethods::Simple_simulation(
  parameters = estimate_result[["estimate_result"]],
  return_format = "SCE",
  seed = 111
)
SCE_result <- simulate_result[["simulate_result"]]
dim(SCE_result)
head(colData(SCE_result))
head(rowData(SCE_result))
```

## Determin the number of cells and genes

In Simple, we can set `nCells` and `nGenes` directly to specify the number of cells and genes to the simulated data.
Here, we simulate a new dataset with 500 cells and 1000 genes:
```{r customed cell and gene number}
simulate_result <- simmethods::Simple_simulation(
  parameters = estimate_result[["estimate_result"]],
  return_format = "list",
  other_prior = list(nCells = 500,
                     nGenes = 1000),
  seed = 111
)
result <- simulate_result[["simulate_result"]][["count_data"]]
dim(result)
```

